---
title: "Final Project Diabetes Report"
author: "Team 6"
format: html
editor: visual
---

# Introduction

This project uses the Pima Indians Diabetes dataset from Kaggle to predict whether a patient has diabetes based on medical measurements.

# Problem Statement

The goal of this analysis is to build and compare classification models to predict diabetes status using variables such as glucose, blood pressure, BMI, insulin, and age. The model performance will be evaluated to select the best predictive model from this dataset.

# Data Description

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(caret)
library(corrplot)
library(rpart)
library(rpart.plot)
library(randomForest)
library(e1071)
library(gbm)
library(pROC)

set.seed(123)
```

```{r}
diabetes <- read.csv("diabetes.csv")
head(diabetes)
dim(diabetes)
names(diabetes)
```

The dataset contains 768 rows and 9 columns. Outcome is the response variable and indicates whether the patient has diabetes; the other 8 are predictor variables.

```{r}
str(diabetes)
```

The data consists of numeric and integer variables.

```{r}
summary(diabetes)

# Gen AI inspired code for Clean summary statistics table
summary_table <- diabetes %>%
  summarise(across(everything(),
                   list(
                     Min = min,
                     Q1 = ~quantile(.x, 0.25),
                     Median = median,
                     Mean = mean,
                     Q3 = ~quantile(.x, 0.75),
                     Max = max
                   ))) %>%
  pivot_longer(everything(),
               names_to = c("Variable", "Statistic"),
               names_sep = "_",
               values_to = "Value") %>%
  pivot_wider(names_from = Statistic,
              values_from = Value) %>%
  mutate(across(where(is.numeric), ~round(.x, 2)))

knitr::kable(summary_table,
             caption = "Summary Statistics for Diabetes Dataset")
```

The summary statistics show us the minimum, median, mean, and maximum values for each variable. The data set has 768 observations of measurements such as glucose, blood pressure, skin thickness, insulin, BMI, diabetes pedigree function, age, and diabetes outcome. Min is zero for all variables besides Age and Diabetes Pedigree function. The age range is 21-81, with a median of 29 and an average of 33.24, indicating that many people in the dataset can be considered young adults. The average outcome variable is 35% of patients have diabetes. The average glucose level is 120.89, which is close to the median, suggesting that high glucose values are bringing the average up. The same is true for BMI, which has a median of 32 and an average of 31.99. Many variables have a minimum of zero, which is unrealistic. It should be either missing or unrecorded data and should be reviewed before preprocessing and training.

```{r}
colSums(is.na(diabetes))
table(diabetes$Outcome)
prop.table(table(diabetes$Outcome))
colSums(diabetes == 0)
```

The dataset does not contain any missing values coded as NaN; however, several variables contain zero values that may not represent valid clinical measurements. The outcome variable includes 500 patients without diabetes and 268 patients with diabetes, indicating a moderate class imbalance that should be considered during model evaluation. Several predictors contain unrealistic zero values: glucose has 5 zero values, blood pressure has 35, skin thickness has 227, insulin has 374, and BMI has 11. Since these measurements should not realistically be zero, they likely represent missing or unrecorded data and should be addressed before model training.

# Data Exploration

```{r}
diabetes %>% mutate(diabetes_status = factor(Outcome, levels = c(0, 1),
            labels = c("No Diabetes", "Diabetes"))) %>%
ggplot(aes(x = diabetes_status)) +
geom_bar(fill = "skyblue") +
labs(title = "Distribution of Diabetes Outcome",
x = "Outcome/Diabetes status",
y = "Number of Patients") +
theme_minimal()
```

The outcome bar chart shows that the dataset contains more patients without diabetes than with diabetes. There are 500 no-diabetes cases and 268 diabetes cases. This means the dataset is somewhat imbalanced, so model evaluation should consider more than accuracy alone.

```{r}

diabetes %>% select(Pregnancies, Glucose, BloodPressure, SkinThickness,
         Insulin, BMI, DiabetesPedigreeFunction, Age) %>%
  pivot_longer(cols = everything(), names_to = "Predictor",
               values_to = "Value") %>%
  
ggplot(aes(x = Value)) +
geom_histogram(bins = 30, fill = "skyblue", color = "black") +
facet_wrap(~ Predictor, scales = "free") +
labs(title = "Predictor Variables Distribution",
x = "Measurement",
y = "Number of Patients") +
theme_minimal()
```

- Age is right-skewed, most patients are younger in their 20s and 30s, and a few older patients around 80 years old.

- Blood pressure values are mostly concentrated between 60 and 80, but several observations have values of zero. These zero values are medically unrealistic and likely represent missing or unrecorded data.

- BMI values are mostly between 20 and 40, with a few zero values that are medically unrealistic.

- Diabetespedigreefunction values are mostly low and right-skewed, with a few high values.

- Most glucose values are between 100 and 150, but some observations have zero values, which are medically unrealistic.

- Insulin has many zero values, which suggest missing values.

- Pregnancies are right-skewed, with most patients having a lower number of pregnancies.

- Skin thickness also has many zero values, which are medically unrealistic.

```{r}
diabetes %>%
  select(Pregnancies, Glucose, BloodPressure, SkinThickness,
         Insulin, BMI, DiabetesPedigreeFunction, Age) %>%
pivot_longer(cols = everything(),names_to = "Predictor",
               values_to = "Measurement") %>%
ggplot(aes(x = Predictor, y = Measurement)) +
geom_boxplot(fill = "skyblue", color = "black") +
facet_wrap(~ Predictor, scales = "free") +
labs(title = "Overall Boxplots of Predictors",
      x = "Predictor Variable",
      y = "Measurement") +
theme_minimal() + theme(axis.text.x = element_blank())

```

Many predictors have outliers so using the median might be better than using the mean for imputation.

```{r}
diabetes %>% mutate(diabetes_status = factor(Outcome,levels = c(0, 1),
             labels = c("No Diabetes", "Diabetes"))) %>%
select(-Outcome) %>%
  pivot_longer(cols = -diabetes_status,names_to = "Predictor",
               values_to = "Value") %>%
ggplot(aes(x = diabetes_status, y = Value, fill = diabetes_status)) +
geom_boxplot() + facet_wrap(~ Predictor, scales = "free") +
labs(title = "Diabetes Outcome by Predictors",
x = "Outcome",
y = "Predictor value") +
theme_minimal() + theme(legend.position = "none")
```

- AGE: Patients with diabetes generally appear older than patients without diabetes. The diabetes group has a higher median age.

- BLOODPRESSURE: Blood pressure looks slightly higher for the diabetes group, but the two box plots are very similar. There does not appear to be a significant difference.

- BMI: BMI appears higher for patients with diabetes but not significantly different. This spread is a lot less for diabetes compared to no diabetes, meaning diabetes BMI are concentrated closer together.

- DIABETESPEDIGREEFUNCTION: The diabetes group has slightly higher values, and outliers have a higher concentration.

- GLUCOSE: shows one of the clearest differences between the two groups. Patients with diabetes generally have higher glucose values.

- INSULIN: Insulin has many low values and several high outliers. The diabetes group has slightly higher insulin values, slightly lower median and the no-diabetes group has more outliers but there is a lot of overlap.

- PREGNANCY: The diabetes group appears to have a higher median number of pregnancies compared to the no-diabetes group, and the diabetes group does not appear to have outliers.

- SKINTHICKNESS: Skin thickness looks similar between the two groups and a lot of overlap but diabetes has some outliers and a slightly larger spread.

```{r}
outcome_correlations <- cor(diabetes %>% select(where(is.numeric))) %>%
  as.data.frame() %>%
  select(Outcome) %>%
  arrange(desc(abs(Outcome)))

outcome_correlations
```

```{r}
diabetes %>% mutate(diabetes_status = factor(Outcome,levels = c(0, 1),
                                labels = c("No Diabetes", "Diabetes"))) %>%
select(Pregnancies, Glucose, BloodPressure, SkinThickness,
         Insulin, BMI, DiabetesPedigreeFunction, Age, diabetes_status) %>%
pivot_longer(cols = -diabetes_status,names_to = "Predictor",
      values_to = "Value") %>%
  
ggplot(aes(x = Value, fill = diabetes_status)) +
geom_density(alpha = 0.4) +
facet_wrap(~ Predictor, scales = "free") +
labs(title = "Density Plots of Predictors by Diabetes",
x = "Measurement",
y = "Density",
fill = "Outcome") + theme_minimal()
```

- The density plots compare the distribution of each predictor for patients with and without diabetes. Glucose shows the most distinction between the diabetes and no diabetes with the diabetes group having higher values. BMI, age, pregnancies and diabetespedigreefunction show some differences between the groups, suggesting they may be useful predictors but still have more overlap in densities. Bloodpressure, insulin, and skinthickness have the most overlap in density between no diabetes and diabetes. Overall the density plots suggest that multiple variables will likely be needed to predict diabetes status.

```{r}
diabetes %>% mutate(diabetes_status = factor(Outcome, levels = c(0, 1),
            labels = c("No Diabetes", "Diabetes"))) %>%
  
ggplot(aes(x = Glucose, y = BMI, color = diabetes_status)) +
geom_point(alpha = 0.6) +
labs(title = "Scatterplot of Glucose and BMI Diabetes Outcome",
     x = "Glucose",
     y = "BMI",
    color = "Diabetes Outcome") + theme_minimal()
```

The scatter plot shows that diabetes cases appear more often at higher glucose values, with diabetes around 130 and above. There is still a lot of overlap between the diabetes and no-diabetes groups especially for BMI but diabetes has a slightly higher BMI. A few points also appear at glucose or BMI values of 0 which needs to be preprocessed.

```{r}
avg_predictors <- diabetes %>% mutate(diabetes_status = factor(Outcome,
                               levels = c(0, 1),
                               labels = c("No Diabetes", "Diabetes"))) %>%
  group_by(diabetes_status) %>%
  summarise(across(c(Pregnancies, Glucose, BloodPressure, SkinThickness,
                     Insulin, BMI, DiabetesPedigreeFunction, Age),
                   mean),
            .groups = "drop") %>%
  pivot_longer(cols = -diabetes_status,
               names_to = "Predictor",
               values_to = "Average")

ggplot(avg_predictors, aes(x = diabetes_status, y = Average, fill = diabetes_status)) +
geom_col(color = "black") +
facet_wrap(~ Predictor, scales = "free") +
labs(title = "Average Health Measurements by Diabetes Status",
       x = "Diabetes Status",
       y = "Average Measurement") + theme_minimal() +
theme(legend.position = "none")

```

The average predictor values show that the diabetes group has higher averages for most variables. Glucose, BMI, and Insulin show one of the clearest differences between the two groups averages.

```{r}

zero_summary <- diabetes %>%
  summarise(across(everything(), ~sum(.x == 0))) %>%
  pivot_longer(cols = everything(),
               names_to = "Variable",
               values_to = "Zero_Count") %>%
  mutate(Zero_Percent = round((Zero_Count / nrow(diabetes)) * 100, 2))


zero_summary %>%
  filter(Variable != "Outcome") %>%
  ggplot(aes(x = reorder(Variable, Zero_Percent), y = Zero_Percent)) +
  geom_col(fill = "skyblue", color = "black") +
  coord_flip() +
  labs(title = "Percentage of Zero Values by Variables",
       x = "Variable",
       y = "Zero Values (%)") +
  theme_minimal()
```

```{r}
count_outliers <- function(x) {
  Q1 <- quantile(x, 0.25)
  Q3 <- quantile(x, 0.75)
  IQR_value <- IQR(x)
  lower_bound <- Q1 - 1.5 * IQR_value
  upper_bound <- Q3 + 1.5 * IQR_value
  sum(x < lower_bound | x > upper_bound)
}

outlier_summary <- diabetes %>%
  select(Pregnancies, Glucose, BloodPressure, SkinThickness,
         Insulin, BMI, DiabetesPedigreeFunction, Age) %>%
  summarise(across(everything(), count_outliers)) %>%
  pivot_longer(cols = everything(),
               names_to = "Variable",
               values_to = "Outlier_Count")
outlier_summary
```

# Data Preprocessing

```{r}
diabetes_clean <- diabetes

zero_na_vars <- c("Glucose", "BloodPressure", "SkinThickness", "Insulin", "BMI")

diabetes_clean[zero_na_vars] <- lapply(diabetes_clean[zero_na_vars], function(x) {
  ifelse(x == 0, NA, x)
})

diabetes_clean$Outcome <- factor(diabetes_clean$Outcome,
                                 levels = c(0, 1),
                                 labels = c("No_Diabetes", "Diabetes"))
```

Unrealistic zero values in glucose, blood pressure, skin thickness, insulin, and BMI were converted to NA because these values are not valid clinical measurements. The response variable, Outcome, was also converted into a factor with two classes: No Diabetes and Diabetes.

```{r}
set.seed(123)

train_index <- createDataPartition(diabetes_clean$Outcome,
                                   p = 0.80,
                                   list = FALSE)

train_data <- diabetes_clean[train_index, ]
test_data <- diabetes_clean[-train_index, ]

train_medians <- sapply(train_data[zero_na_vars], median, na.rm = TRUE)
```

The cleaned dataset was split into training and testing sets using an 80/20 split. The training set will be used to build the models, while the testing set will be used to evaluate model performance on unseen data.

After the split, median values were calculated using only the training data. These medians will be used to fill the missing values in both the training and testing sets. Calculating the medians from the training set helps avoid data leakage because the test set does not influence the preprocessing values.

```{r}
for (var in zero_na_vars) {
  train_data[[var]][is.na(train_data[[var]])] <- train_medians[var]
  test_data[[var]][is.na(test_data[[var]])] <- train_medians[var]
}

colSums(is.na(train_data))
colSums(is.na(test_data))


dim(train_data)
dim(test_data)

prop.table(table(train_data$Outcome))
prop.table(table(test_data$Outcome))
```

Missing values in the training and testing sets were filled using median values calculated from the training set. After imputation both sets had 0 missing values.

The final split produced 615 training observations and 153 testing observations.

```{r}

train_medians
```

The training medians show the values used to replace missing measurements in both the training and testing sets. These values were calculated from the training data only to avoid data leakage.

```{r}
numeric_predictors <- train_data %>% 
  mutate(Outcome = as.numeric(Outcome)) %>%
  select(where(is.numeric))

cor_matrix <- cor(numeric_predictors)

corrplot(cor_matrix, method = "color",
        addCoef.col = "black",
        number.cex = 0.60,
        tl.cex = 0.8)
```

The correlation plot shows that glucose has the strongest positive relationship with outcome at a correlation of 0.49. This suggests that higher glucose values are associated with a higher chance of having diabetes. BMI, Age, Pregnancies and DiabetesPedigreeFunction also have positive correlations with the outcome variable, but these relationships are weaker than the correlation between glucose and outcome. The correlation plot also shows moderate positive relationships among some predictors. BMI and SkinThickness have a correlation of 0.56, while Insulin and Glucose have a correlation of 0.42. Overall Glucose, BMI, Age, Pregnancies and DiabetesPedigreeFunction may be useful predictors.

# Modeling

Scale features for KNN and SVN

```{r}
# Features to scale 
scale_vars <- c("Pregnancies", "Glucose", "BloodPressure", "SkinThickness",
                "Insulin", "BMI", "DiabetesPedigreeFunction", "Age")
 
train_scaled <- train_data
test_scaled  <- test_data
 
# Scale using TRAINING means and SDs only (no leakage)
train_means <- colMeans(train_data[scale_vars])
train_sds   <- apply(train_data[scale_vars], 2, sd)
 
train_scaled[scale_vars] <- scale(train_data[scale_vars])
test_scaled[scale_vars]  <- scale(test_data[scale_vars],
                                  center = train_means,
                                  scale  = train_sds)
```

Since KNN and SVM are distance-based algorithms, all features were standardised using the mean and standard deviation calculated from the training set only to avoid data leakage.

Extract metrics from confusion matrix

```{r}
get_metrics <- function(pred, actual, model_name, prob = NULL) {
  cm <- confusionMatrix(pred, actual, positive = "Diabetes")
  auc_val <- if (!is.null(prob)) {
    roc_obj <- roc(actual, prob, levels = c("No_Diabetes", "Diabetes"), quiet = TRUE)
    round(auc(roc_obj), 4)
  } else NA
 
  data.frame(
    Model       = model_name,
    Accuracy    = round(cm$overall["Accuracy"], 4),
    Sensitivity = round(cm$byClass["Sensitivity"], 4),
    Specificity = round(cm$byClass["Specificity"], 4),
    F1          = round(cm$byClass["F1"], 4),
    AUC         = auc_val
  )
}

```

## Logistic Regression

```{r}
set.seed(123)
log_model <- glm(Outcome ~ ., data = train_data, family = binomial)
summary(log_model)
 
log_prob <- predict(log_model, newdata = test_data, type = "response")
log_pred <- factor(ifelse(log_prob >= 0.5, "Diabetes", "No_Diabetes"),
                   levels = c("No_Diabetes", "Diabetes"))
 
cm_log <- confusionMatrix(log_pred, test_data$Outcome, positive = "Diabetes")
cm_log

```

## Decision Tree

```{r}
set.seed(123)
tree_model <- rpart(Outcome ~ ., data = train_data, method = "class",
                    control = rpart.control(cp = 0.01))
rpart.plot(tree_model, main = "Decision Tree")
 
tree_prob <- predict(tree_model, newdata = test_data, type = "prob")[, "Diabetes"]
tree_pred <- predict(tree_model, newdata = test_data, type = "class")
 
cm_tree <- confusionMatrix(tree_pred, test_data$Outcome, positive = "Diabetes")
cm_tree

```

## Random Forest

```{r}
set.seed(123)
rf_model <- randomForest(Outcome ~ ., data = train_data,
                         ntree = 500, importance = TRUE)
rf_model
 
rf_prob <- predict(rf_model, newdata = test_data, type = "prob")[, "Diabetes"]
rf_pred <- predict(rf_model, newdata = test_data, type = "class")
 
cm_rf <- confusionMatrix(rf_pred, test_data$Outcome, positive = "Diabetes")
cm_rf
 
# Variable importance plot
varImpPlot(rf_model, main = "Random Forest Variable Importance")

```

## KNN

```{r}
set.seed(123)
 
# Use cross-validation to find the best k
ctrl <- trainControl(method = "cv", number = 5,
                     classProbs = TRUE, summaryFunction = twoClassSummary)
 
knn_cv <- train(Outcome ~ ., data = train_scaled,
                method = "knn",
                trControl = ctrl,
                metric = "ROC",
                tuneGrid = data.frame(k = seq(3, 21, by = 2)))
 
best_k <- knn_cv$bestTune$k
cat("Best k:", best_k, "\n")
plot(knn_cv, main = "KNN Cross-Validation: k vs AUC")
 
knn_prob <- predict(knn_cv, newdata = test_scaled, type = "prob")[, "Diabetes"]
knn_pred <- predict(knn_cv, newdata = test_scaled)
 
cm_knn <- confusionMatrix(knn_pred, test_scaled$Outcome, positive = "Diabetes")
cm_knn

```

## SVM

```{r}
set.seed(123)
svm_model <- svm(Outcome ~ ., data = train_scaled,
                 kernel = "radial", probability = TRUE, cost = 1)
 
svm_raw  <- predict(svm_model, newdata = test_scaled, probability = TRUE)
svm_pred <- svm_raw
svm_prob <- attr(svm_raw, "probabilities")[, "Diabetes"]
 
cm_svm <- confusionMatrix(svm_pred, test_scaled$Outcome, positive = "Diabetes")
cm_svm
```

## Gradient Boosting

```{r}
set.seed(123)
 
# gbm needs numeric outcome
train_gbm <- train_data
test_gbm  <- test_data
train_gbm$Outcome_num <- ifelse(train_gbm$Outcome == "Diabetes", 1, 0)
train_gbm$Outcome     <- NULL
 
gbm_model <- gbm(Outcome_num ~ ., data = train_gbm,
                 distribution = "bernoulli",
                 n.trees = 500, interaction.depth = 3,
                 shrinkage = 0.01, cv.folds = 5, verbose = FALSE)
 
best_trees <- gbm.perf(gbm_model, method = "cv", plot.it = TRUE)
cat("Best number of trees:", best_trees, "\n")
 
gbm_prob <- predict(gbm_model, newdata = test_gbm,
                    n.trees = best_trees, type = "response")
gbm_pred <- factor(ifelse(gbm_prob >= 0.5, "Diabetes", "No_Diabetes"),
                   levels = c("No_Diabetes", "Diabetes"))
 
cm_gbm <- confusionMatrix(gbm_pred, test_data$Outcome, positive = "Diabetes")
cm_gbm

```

## Model Comparsion Table

```{r}
results <- rbind(
  get_metrics(log_pred,  test_data$Outcome,        "Logistic Regression", log_prob),
  get_metrics(tree_pred, test_data$Outcome,        "Decision Tree",       tree_prob),
  get_metrics(rf_pred,   test_data$Outcome,        "Random Forest",       rf_prob),
  get_metrics(knn_pred,  test_scaled$Outcome,      "KNN",                 knn_prob),
  get_metrics(svm_pred,  test_scaled$Outcome,      "SVM",                 svm_prob),
  get_metrics(gbm_pred,  test_data$Outcome,        "Gradient Boosting",   gbm_prob)
)
 
rownames(results) <- NULL
 
knitr::kable(results, caption = "Model Performance Comparison")
```

```{r}
roc_log  <- roc(test_data$Outcome,    log_prob,  levels = c("No_Diabetes", "Diabetes"), quiet = TRUE)
roc_tree <- roc(test_data$Outcome,    tree_prob, levels = c("No_Diabetes", "Diabetes"), quiet = TRUE)
roc_rf   <- roc(test_data$Outcome,    rf_prob,   levels = c("No_Diabetes", "Diabetes"), quiet = TRUE)
roc_knn  <- roc(test_scaled$Outcome,  knn_prob,  levels = c("No_Diabetes", "Diabetes"), quiet = TRUE)
roc_svm  <- roc(test_scaled$Outcome,  svm_prob,  levels = c("No_Diabetes", "Diabetes"), quiet = TRUE)
roc_gbm  <- roc(test_data$Outcome,    gbm_prob,  levels = c("No_Diabetes", "Diabetes"), quiet = TRUE)

ggroc(list(
  "Logistic Regression" = roc_log,
  "Decision Tree"       = roc_tree,
  "Random Forest"       = roc_rf,
  "KNN"                 = roc_knn,
  "SVM"                 = roc_svm,
  "Gradient Boosting"   = roc_gbm
), legacy.axes = TRUE, size = 0.9) +
  geom_abline(linetype = "dashed", color = "grey50") +
  labs(title = "ROC Curves for All Models",
       x = "1 - Specificity", y = "Sensitivity", color = "Model") +
  theme_minimal()

```

